---
title: "Forest Fires in Portugal"
author: João Teixeira - up200705307, Nuno Peixoto - up200804621
date: "25-01-2015"
output:
  beamer_presentation:
    toc: true
theme: Antibes
---

# Clustering

The following clustering methods were used to try to find different groups of observations present in the data set:

* Clustering Large Applications (CLARA)
* Partitioning Around Medoids (PAM)
* Hierarchical Clustering
* K-Means Clustering

## Clustering results

Using a R script with the help of the silhouette function we could find the best number of clusters for each used method 

| CLARA   |         |PAM      |         |
|---------|---------|---------|---------|
|nClusters|SilhCo   |nClusters|SilhCo   |
|2        |0.7144933|2        |0.7336859|
|4        |0.6361579|3        |0.6187632|
|3        |0.6243904|4        |0.4812423|
|5        |0.3102661|5        |0.4318594|
|10       |0.2789919|6        |0.3595906|
|9        |0.2749504|7        |0.3234947|
|6        |0.2645142|9        |0.2750479|
|7        |0.2329236|8        |0.2561567|
|8        |0.2286001|10       |0.2560370|

## Clustering results (cont.)

|hclust   |          |kmeans   |         |
|---------|----------|---------|---------|
|nClusters|SilhCo    |nClusters|SilhCo   | 
|2        |0.55027091|2        |0.7791398|
|3        |0.32391734|3        |0.6993844|
|5        |0.17051099|4        |0.6888222|
|4        |0.16893921|5        |0.5839630|
|6        |0.09489308|6        |0.5492905|
|7        |0.09160570|7        |0.3886676|
|8        |0.08740959|8        |0.3868419|
|10       |0.08135908|10       |0.3321750|
|9        |0.07454503|9        |0.3175380|

## Silhouette Plot

```{r, echo=FALSE}
k2 = kmeans(new_train_data[,-73], centers=2, iter.max=300)
s = silhouette(k2$cluster,dist(new_train_data[,-73]))
plot(s)
```

# Predictive Models

In order to find the best regression model that can predict the target variable of the test data set with less error, we analysed the following forecasting models:

* Multiple Linear Regression
* Regression Trees
* K-Nearest Neighbors (KNNs)
* Support Vector Machines (SVMs)
* Artificial Neural Networks (ANNs)
* Random Forest (Ensembles)

## Model comparison

The metric evaluation to be considered for these forecasts is the MAE - Mean Absolute Error. We use a
cross validation method with 2 repetitions of 3 folds.

To evaluate the models we will use the performanceEstimation package that provides a set of functions
and arguments that allow us to change the values of parameters in order to check the best fit for an specific
model.

## Model performance

| **Model**                        | **MAE ERROR** | **Parameters**                                                                           |
| -------------------------------- |:-------------:| -----------------------------------------------------------------------------------------|
| SVM (ksvm)                       |1732.284       |  **epsilon**:1e-09, **C**:1, **kernel**:rbfdot                                           |
| SVM                              |1762.223       |  **cost**:1, **gamma**:0.01                                                              |
| k-Nearest Neighbors              |1916.375       |   **scale**:TRUE, **k**:11, **distance**:1, **kernel**:triangular                        |
| Multiple Linear Regression       |2310.785       |   Default Parameters                                                                     |
| ANN                              |2393.626       |   **size**:2, **maxit**:200, **decay**:0.1, **scale**:TRUE, **trace**:FALSE, **linout**:1|
| Random Forest                    |2393.626       | **ntree**:500, **nodesize**:5, **corr.bias**:FALSE, **mtry**:3                           |
| Regression Trees                 |2802.628       |   **se**:1, **minsplit**:15                                                              |

## Models Plot

```{r total, echo=FALSE, warning=FALSE, comment=FALSE, comment=FALSE, message=FALSE, results='hide', dpi=100}
total <- performanceEstimation(
  PredTask(TotalBurntArea ~ ., new_train_data),
  c(workflowVariants(learner="lm", pre="centralImp",post="onlyPos"),
    workflowVariants(learner="rpartXse",
                     learner.pars=list(se=1, minsplit=20)),
    workflowVariants(learner="train.kknn",
                     learner.pars=list(scale=T, k=13, distance=1, kernel="epanechnikov")),
    workflowVariants(learner="svm",
                     learner.pars=list(cost=1, gamma=0.01)),
    workflowVariants(learner="ksvm",
                     learner.pars=list(epsilon=10^-9, C=1, kernel="rbfdot")),
    workflowVariants(learner="nnet",
                     learner.pars=list(size=4, maxit=300, decay=0.4, scale=T, trace=F, linout=1)),
    workflowVariants(learner="randomForest",
                     learner.pars=list(ntree=250, nodesize=5, corr.bias=F, mtry=9))),
  EstimationTask(metrics="mae",method=CV(nReps=1,nFolds=2)))

plot(total, ylab="Model", xlab="MAE Results")
```

# Conclusion

**Best performance models**
  * **SVM** (ksvm)
    * MAE error: 1732.284
    
  * **SVM**
    * MAE error: 1762.223
    
  * **k-Nearest Neighbors**
    * MAE error: 1916.375


